---
bibliography: references.bib
---

`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'`

```{=html}
<!--
If you feel it necessary to include an appendix, it goes here.
-->
```

# Additional Data {#data}

## Tree Height and Crown Width Predictive Model Equations {#tree-model-eqs} 

\textbf{ACMA equations:} (\textit{x} = DBH) 
\begin{equation}
\textrm{Tree Height} = 8.977 + 3.617x -0.049x^{2} + 0.001x^{3}  \\
\end{equation}

\begin{equation}
\textrm{Crown Width} = 13.01 + 1.593x - 0.007x^{2} + 0.0004x^{3} \\
\end{equation}

\textbf{ACPL equations:} (\textit{x} = DBH) 
\begin{equation}
\textrm{Tree Height} = 10.955+ 2.67x -0.032x^2 + 0.001x^3  \\
\end{equation}

\begin{equation}
\textrm{Crown Width} = 6.445 + 2.29x - 0.009x^2 - 0.0002x^3 \\
\end{equation}

\textbf{PSME equations:} (\textit{x} = DBH) 
\begin{equation}
\textrm{Tree Height} = -4.497 + 7.516x - 0.146x^2 + 0.001x^3 \\
\end{equation}

\begin{equation}
\textrm{Crown Width} = 6.536 + 1.471x - 0.013x^2 + 0.0005x^3 \\
\end{equation}

\textbf{THPL equations:} (\textit{x} = DBH) 
\begin{equation}
\textrm{Tree Height} = 7.311 + 3.457x - 0.0473x^2 + 0.001x^3 \\
\end{equation}

\begin{equation}
\textrm{Crown Width} = 5.113 + 1.622x - 0.032x^2 + 0.0002x^3 \\
\end{equation}


# Code {#code-chunks}


This second appendix includes all of the R chunks of code that were hidden throughout the document to help with readability and/or setup.

## **In Chapter** \@ref(data-methods)**:**

### Portland Tree Inventory counts and calculations {-}

```{r ref.label='counttables', results='hide', echo = TRUE}
```

### Calculating NDVI from satellite imagery {-}

```{r ref.label='ndvi-calc', results='hide', echo = TRUE, eval = F}
```

### LiDAR canopy delineation {-}

```{r ref.label='lidar-calc', results='hide', echo = TRUE, eval = F}
```

## **In Chapter** \@ref(results)**:**
